Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-31 bn.tex
Typos

Bayesian learning of Bayesian network structures is based on the
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyper\nobreakdash-parameters for the model parameters $\theta$, and
the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
vectors collected to a $N\times n$ design matrix.  We use structure
$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
to denote those rows of the column $i$ in which the parents $G_i$
contain the value configuration number $j\in\{1,\ldots,q_i\}$.

@@

Finding satisfactory Dirichlet hyperparameters for the Bayesian
mixture above has, however, turned out to be problematic. Early on,
one of the desiderata for a good model selection criterion was that it
is score-equivalent, i.e., it would yield equal scores for
essentially equivalent models~\cite{Verm90}.  For example, the score
for the structure $X_1\rightarrow X_2$ should be the same as the score
for the model $X_2 \rightarrow X_1$ since they both correspond to the
hypothesis that variables $X_1$ and $X_2$ are statistically dependent
on each other.  It can be shown~\cite{Heck95} that to achieve this,
not all the hyperparameters $\alpha$ are possible and for practical
reasons Buntine~\cite{Bunt91} suggested a so called BDeu score with
just one hyperparameter $\alpha\in R_{++}$ so that
$\theta_{ij\cdot}\sim Dir(\frac{\alpha}{q_i
  r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However, it soon turned out
that the BDeu score was very sensitive to the selection of this
hyperparameter~\cite{cosco.uai07} and that for small sample sizes this
method detects spurious correlations~\cite{Steck08} leading to models
with suspiciously many parameters.

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score-equivalent.

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. This information theoretic NML criterion
can be justified from the minimum description length point of view,
\cite{Riss78,Grun07}. It has been shown to be robust with respect to
different data generating mechanisms where a good choice of prior
is challenging, see~\cite{eggeling2014robust,maatta16}. While it is
easy to see that the NML criterion satisfies the requirement of giving
equal scores to equal structures, the normalizing constant renders the
computation infeasible.

>>>>

Bayesian learning of Bayesian network structures is based on the
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyperparameters for the model parameters $\theta$, and
the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
vectors collected to a $N\times n$ design matrix.  We use structure
$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
to denote those rows of the column $i$ in which the parents $G_i$
contain the value configuration number $j\in\{1,\ldots,q_i\}$.

@@

Finding satisfactory Dirichlet hyperparameters for the Bayesian
mixture above has, however, turned out to be problematic. Early on,
one of the desiderata for a good model selection criterion was that it
is score equivalent, i.e., it would yield equal scores for
essentially equivalent models~\cite{Verm90}.  For example, the score
for the structure $X_1\rightarrow X_2$ should be the same as the score
for the model $X_2 \rightarrow X_1$ since they both correspond to the
hypothesis that variables $X_1$ and $X_2$ are statistically dependent
on each other.  It can be shown~\cite{Heck95} that to achieve this,
not all the hyperparameters $\alpha$ are possible and for practical
reasons Buntine~\cite{Bunt91} suggested a so-called BDeu score with
just one hyperparameter $\alpha\in R_{++}$ so that
$\theta_{ij\cdot}\sim Dir(\frac{\alpha}{q_i
  r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However, it soon turned out
that the BDeu score was very sensitive to the selection of this
hyperparameter~\cite{cosco.uai07} and that for small sample sizes this
method detects spurious correlations~\cite{Steck08} leading to models
with suspiciously many parameters.

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent.

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. This information-theoretic NML criterion
can be justified from the minimum description length point of view,
\cite{Riss78,Grun07}. It has been shown to be robust with respect to
different data generating mechanisms where a good choice of prior
is challenging, see~\cite{eggeling2014robust,maatta16}. While it is
easy to see that the NML criterion satisfies the requirement of giving
equal scores to equal structures, the normalizing constant renders the
computation infeasible.

### 2017-03-31 experiments.tex
finetuning
Typos

@@

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
\caption{Finding generating models of 4 arcs.}
\label{fig:4arcs}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
\caption{Finding generating models of 7 arcs.}
\label{fig:4arcs}
\end{figure}

@@

We first compare the ability of the different scoring criteria to
discover the structure that generated the data. For this purpose we
create 100 random 5-node Bayesian network structures with 4
edges and another 100 structures with 7 edges.  The variables were
randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
each network, we generated parameters by two different schemes. The
first scheme exactly matched the assumptions of the BDeu score with
$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
was to generate the parameters independently from a Dirichlet
distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
that is the Jeffreys' prior for the multinomial model. This distribution
was selected instead of the uniform distribution in order to make the
generating structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, BDeu usually
also ranks the generating structure higher than the other scores
(Figure X).  On the other hand, when the parameters are drawn from the
Jeffreys' prior, the fNML score appears to rank the generating structure
highest. This too is not surprising, since the NML distribution for
multinomial data is known to closely match the marginal distribution of
the data when the Jefrreys' prior is used.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
dataset was selected as training data and the rest 90\% as test data.
The training was conducted using a dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

When predicting with structures learned by the BDeu score, we used the
Bayesian expected parameter values $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter free, for structures learned by the other model
selection criteria, we used the predictive conditional NML
parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
\caption{Number of parameters in a breast cancer model as a function
  of sample size for different model selection criteria.}
\label{fig:bcnpmean}
\end{figure}

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
\caption{Finding generating models of 4 arcs.}
\label{fig:4arcs}
\end{figure}

@@

@@

We first compare the ability of the different scoring criteria to
discover the structure that generated the data. For this purpose we
create 100 random 5-node Bayesian network structures with 4
edges and another 100 structures with 7 edges.  The variables were
randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
each network, we generated parameters by two different schemes. The
first scheme exactly matched the assumptions of the BDeu score with
$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
was to generate the parameters independently from a Dirichlet
distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$,
that is, the Jeffreys' prior for the multinomial model. This distribution
was selected instead of the uniform distribution in order to make the
generating structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

\begin{figure}[h]
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
\caption{Finding generating models of 7 arcs.}
\label{fig:4arcs}
\end{figure}

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, BDeu usually
also ranks the generating structure higher than the other scores
(Figure X).  On the other hand, when the parameters are drawn from the
Jeffreys' prior, the fNML score appears to rank the generating structure
highest. This too is not surprising, since the NML distribution for
multinomial data is known to closely match the marginal distribution of
the data when the Jeffreys' prior is used.

@@

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
dataset was selected as training data and the rest 90\% as test data.
The training was conducted using dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

When predicting with structures learned by the BDeu score, we used the
Bayesian expected parameter values $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter-free, for structures learned by the other model
selection criteria, we used the predictive conditional NML
parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

### 2017-03-31 intro.tex
Typos

Learning the structure of statistical dependencies can be seen as a
model-selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
identifying the appropriate structure and/or predictive
performance~\cite{cosco.pgm08a}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but studies~\cite{cosco.uai07,Steck08}
have found this criterion to be very sensitive to hyper parameters
and to yield undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

>>>>

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
identifying the appropriate structure and/or predictive
performance~\cite{cosco.pgm08a}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but studies~\cite{cosco.uai07,Steck08}
have found this criterion to be very sensitive to hyperparameters
and to yield undesirably complex models for small sample sizes.

The information-theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

### 2017-03-31 qnml.tex
Typos

qNML yields equal scores for network structures that encode same sets
of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those which a) are same when directed
arcs are substituted by undirected ones and b) which have same
\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
that all the equivalent network structures, and only those structures,
can be reached from each other by reversing, one by one, the so called
\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
$B$'s parents other than $A$ are exactly  $A$'s parents
($G_B=\{A\}\cup G_A$).

@@

One important property possessed by nearly every model selection
criterion is consistency. In our context, consistency means that given
a data matrix with $N$ samples coming from a distribution faithful to
some DAG $G$, the qNML will give the highest score to true graph $G$
with a probability tending to one as $N$ increases. We will show this
by first proving that qNML is asymptotically equivalent to widely used
BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
The outline of this proof follows a similar pattern to that in
\cite{SilanderIJAR10} where the consistency of fNML was proved.

>>>>

qNML yields equal scores for network structures that encode same sets
of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those which a) are same when directed
arcs are substituted by undirected ones and b) which have same
\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
that all the equivalent network structures, and only those structures,
can be reached from each other by reversing, one by one, the so-called
\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
$B$'s parents other than $A$ are exactly  $A$'s parents
($G_B=\{A\}\cup G_A$).

@@

One important property possessed by nearly every model selection
criterion is consistency. In our context, consistency means that given
a data matrix with $N$ samples coming from a distribution faithful to
some DAG $G$, the qNML will give the highest score to true graph $G$
with a probability tending to one as $N$ increases. We will show this
by first proving that qNML is asymptotically equivalent to the widely used
BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
The outline of this proof follows a similar pattern to that in
\cite{SilanderIJAR10} where the consistency of fNML was proved.
